The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-06 thesis/extra_results.tex
thesis; julio oct 4 comments

\chapter{Additional results an tests}\label{sec:extra_results}

In this section, we briefly explore additional simulations testing
assumptions not varied in the main text: variations in the halo
concentration, cored halos, alternative orbits, anisotropy, and
ellipsoidal models.

@@

The initial concentration (i.e.~choice of \(\vmax\) and \(\rmax\))
affects total dark matter evolution.
Fig.~\ref{fig:tidal_tracks_concentration} compares models on the
\smallperi{} orbit with different initial halo parameters. The heavier
halo has \(\vmax=43\) and \(\rmax=7\), whereas the lighter halo has
\(\vmax=25\) and \(\rmax=2.5\). While the less concentrated halo loses
more mass, the halo evolves to similar final structural parameters, as
expected in \citet{EN2021}. We note these choices were constrained to
match Scl's present-day velocity dispersion.

@@

We have good observational evidence for some dwarf galaxies to be cored.
Even Sculptor has debated claims for having a core
\citep[e.g.,][]{amorisco+zavala+deboer2014, breddels+helmi2013, battaglia+2008, walker+2009a, richardson+fairbairn2014, agnello+evans2012}.
To test if a core matter, we adopt a ``cored-NFW'' model,
\begin{equation}{
\rho/\rho   _s = \frac{1}{(1+r/r_s)^2 (r_c/r_s + r/r_s)},
}\end{equation} where \(r_c\) is the core radius. We adopt
\(r_c = 0.1r_s\) and adopt an extreme initial mass of \(\vmax=49\,\kms\)
and \(\rmax=7.1\,\kpc\).

Sculptor and Ursa Minor may also be highly elliptical in 3D
\citep[e.g.,][]{an+koposov2022}, especially given their projected
shapes. We consider

We find the evolution of the oblate halo to be nearly identical to the
spherical halo, similar to \citet{battaglia+sollima+nipoti2015}.

While we have compared the evolution of orbits in the MW potential for
simplicity, because the tidal evolution of Scl is very weak in the LMC
model, the differences between models become even more slight.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_orbits_boundmass.pdf}}
\caption[Ursa Minor tidal dependence on orbit]{Similar to
Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
orbits in the LMC. Most LMC models evolve more weakly than the MW
models, however these do not reach full agreement with the present-day
position so misrepresent recent tidal
evolution.}\label{fig:tidal_tracks_umi}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_impact_i_f.pdf}}
\caption[Scl MW impact stellar densities]{The initial and final density
profiles for the Sculptor model which passes through the MW previously.
The impact does not substantially affect the inner density region. The
hints of a tidal stream in this model are misaligned to the proper
motion, likely a result of the LMC.}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_stream.pdf}}
\caption[Scl predicted stream]{The predicted properties of a tidal tail
in the Scl model.}\label{fig:scl_tidal_stream}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_impact_stream.pdf}}
\caption[Scl predicted stream]{the predicted properties of the stream of
stars around Scl in the MW impact model. Note that, compared to
Fig.~\ref{fig:scl_sim_stream}, this model produces a much more scattered
and misaligned stream, due to the long time since stream formation and
nonlinear effects of the LMC encounter. \textbf{todo check 2d velocity
gradients in this case}}\label{fig:scl_mw_impact_tidal_stream}
\end{figure}

>>>>

\chapter{Additional simulation tests}\label{sec:extra_results}

In this section, we briefly explore additional simulations testing
assumptions not varied in the main text: variations in the halo
concentration, cored halos, alternative orbits, anisotropy, and
ellipsoidal models. While each of these variables influences resulting
dark matter evolution, the final stellar component and dark matter
remnant remains largely across alternate initial conditions. We conclude
that consideration of these effects likely would not change our
conclusions.

The following models aim to reproduce the velocity dispersion
\(\lesssim 1\,\kms\) with a similar present-day half-light radius.

@@

\subsection{Halo concentration}\label{halo-concentration}

Changing the initial concentration (or \(\vmax\) and \(\rmax\))
primarily affects total dark matter evolution.
Fig.~\ref{fig:tidal_tracks_concentration} compares models on the
\smallperi{} orbit with different initial halo parameters. The heavier
halo has \(\vmax=43\) and \(\rmax=7\), whereas the lighter halo has
\(\vmax=25\) and \(\rmax=2.5\). While the less concentrated halo loses
more mass, the halo evolves to similar final structural parameters.

@@

\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}

Figure: The evolution of velocity anisotropy.

\subsection{Dark matter cores}\label{dark-matter-cores}

Many dwarf galaxies appear to have dark matter cores. In Scl, the
presence or absence of a core has been debated
\citep[e.g.,][]{battaglia+2008, walker+2009a, agnello+evans2012, breddels+helmi2013, amorisco+zavala+deboer2014, richardson+fairbairn2014}.
To simulate the evolution of a cored dark matter halo, we adopt a
``cored-NFW'' model, \begin{equation}{
\rho/\rho   _s = \frac{1}{(1+r/r_s)^2 (r_c/r_s + r/r_s)},
}\end{equation} where \(r_c\) is the core radius. We set
\(M_s=0.79\times10^{10}\,\Mo\) where \(M_s = 4\pi/3\ r_s^3 \rho_s\),
\(r_s=2.76\,\kpc\), \(r_c=0.1r_s\), and use the same truncation as our
fiducial halo (Eq.~\ref{eq:trunc_nfw}). While this is not the largest
possible core, the core is similar in size to the stellar component.
Larger cores would require a more massive halo in order to match the
velocity dispersion at a constant scale radius.

In Fig.~\ref{fig:tidal_tracks_structure}, we show the evolution of the
cored model as compared to a heavier halo. The cored halo loses more
total dark matter mass, evolving further along the tidal track as
expected. However, in order to reproduce the observed velocity
dispersion, the final halo reaches a \(\rmax\) and \(\vmax\) as the
fiducial halo. A cored dark matter halo likely cannot be completely
tidally disrupted and still recover the observed velocity dispersion.

\pandocbounded{\includegraphics[keepaspectratio]{figures/cored_density_i_f.pdf}}

Figure: Cored versus cuspy initial and final density (todo, fix)..

\subsection{An ellipsoidal halo}\label{an-ellipsoidal-halo}

Sculptor and Ursa Minor may be highly elliptical in 3D, possibly
violating our assumption of spherical symmetry
\citep[e.g.,][]{an+koposov2022}.

We create an oblate initial dark matter halo using \agama{}. The initial
snapshot is sampled from the distribution function \begin{equation}{
\begin{split}
f = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h(J)}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g(J)}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g(J)}{J_{\rm cutoff}} \right)^\zeta\right] 
\end{split},
}\end{equation} (\texttt{doublePowerLaw} in \agama{}), with
\(g(J) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
\(h(J) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
\texttt{example\_doublepowerlaw.py} script in \Agama{} solves for the
best parameters matching a given density profile. We chose to create a
model resembling an NFW but scaled by a factor 0.5 in the \(z\) axis.
For a scale-free halo (\(r_s=1\), \(M_s=1\)), the best-fit parameters
are \(J_0=0.890\), \(\Gamma=1.46\), \(\eta=0.568\), \(B=2.97\),
\(h_r=0.845\), \(h_z=1.66\), \(g_r=0.753\), \(g_z=1.69\), \(M_0=0.965\),
\(J_{\rm cutoff}=2.40\), \(\zeta=20\). The resulting initial conditions
are stable in isolation and produce a halo with near the desired density
profile and ellipticity. We scale the halo to have a major axis density
profile equivalent to an NFW with \(\rmax = 7\,\kpc\) and
\(\vmax=48\,\kpc\).

We find the evolution of the oblate halo to be nearly identical to the
spherical halo, similar to \citet{battaglia+sollima+nipoti2015}. The
oblate halo however becomes spherical as time progresses. If Scl's dark
matter halo is indeed elliptical, this model may not be an adequate
description. Possibly, internal rotation, alternative density structure,
or velocity anisotropy play a role in the stability of oblate initial
conditions to tidal stripping. A full investigation of these effects is
left to future work.

\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_density_i_f.pdf}}

Slices of the oblate density before and after tidal evolution.
\textbf{TODO: isodensity contours on RHS projected major/minor axis}.

@@

\subsection{Orbital variation}\label{orbital-variation}

In the main text, we focus on recent tidal impacts on Scl and UMi. But,
due to the long-term orbital uncertainties of Scl and UMi with an LMC
(see Fig.~\ref{fig:scl_umi_orbit_uncert}), tidally extreme orbits are
entirely possible. Here, we consider a few examples of extreme orbital
evolution in the long term case. While the actual orbit of either galaxy
almost certainly deviates from these examples (discussion in
Section~\ref{sec:orbital_uncertainties}), these orbits still represent
extreme past tidal histories.

For a more massive LMC (e.g., the L3M11 model), Scl may have undergone
an extreme pericentric passage with the MW \(\sim 6\,\Gyr\) ago. We find
that the 3\(\sigma\) smallest pericentre is \(4\,\kpc\). To test if this
model forms a different density profile, we simulate this model.

Because of the strong tidal interaction with the MW, the trajectory is
substantially perturbed from a point orbit. We adjust the initial
conditions by comparing the change in actions (as calculated in the
static MW-only potential) before and after the pericentre.
Table~\ref{tbl:orbit_adjustments} records the iterations we take. Our
final model is able to approximately reproduce the observed position of
Scl, but more refinement is required for a closer match.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_orbits_boundmass.pdf}}
\caption[Ursa Minor tidal dependence on orbit]{Similar to
Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
orbits in the LMC. Most LMC models evolve more weakly than the MW
models, however these do not reach full agreement with the present-day
position so misrepresent recent tidal
evolution.}\label{fig:tidal_tracks_umi}
\end{figure}

\section{Present-day properties}\label{present-day-properties}

We find that this model also does not display strong signs of tidal
signatures. Fig.~\ref{fig:scl_mw_impact_i_f} shows the initial and final
stellar profiles for this model (integrated with a lower-resolution
\(10^6\) particles). The model evolves little more than the
\smallperi{}. Given that the Jacobi radius would be within \(R_h\), why
does this model not show strong density perturbations?

We propose that th case Scl undergoes a small MW pericentre is an
impulsive encounter. Because Scl's apocentre is still \(\sim 300\,\kpc\)
in this model, the galaxy passes through the MW with velocity
\(600\,\kms?\). Since the crossing time at the half-light radius is
about 90 Myr, in one crossing-time, the model moves from pericentre to a
Galactocentric distance of \(\sim20\,\kpc\), where tides forced have
dropped substantially. As a result, after \(6\Gyr\) of relatively
isolated evolution, the break radius falls well outside the observed
density profile, but the injection of energy uniformly throughout the
halo only results in moderate DM loss predominantly in the outskirts,
not a reshaping of the inner density profile. Note that the Jacobi
radius is derived on circular orbits, not highly elliptical orbits like
this case.

However, we only note that the stars appear to be consistent but are
well below the \citet{power+2003}'s numerically converged radius.

\section{The formation of tidal
tails}\label{the-formation-of-tidal-tails}

As we predict that any tidal tails formed in Scl and UMi would be
extremely faint, we deferred discussion of the properties of any such
tails to this section. We note that in extreme cases, \(\sim 2\%\) of
stars may become unbound in either galaxy, forming tidal tails which may
be detectible with substantially better observations.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_stream.pdf}}
\caption[Scl predicted stream]{The predicted properties of a tidal tail
in the Scl model. \textbf{TODO: titles}}\label{fig:scl_tidal_stream}
\end{figure}

\section{Summary}\label{summary}

While we have compared the evolution of orbits in the MW potential for
simplicity, because the tidal evolution of Scl is very weak in the LMC
model, the differences between models become even more slight. And while
UMi has a lower pericentre than Scl, the differences between halo
structure should apply similarly to UMi.

### 2025-10-06 thesis/extra_rv_models.tex
thesis; julio oct 4 comments

\chapter{LOS velocity modeling}\label{sec:rv_obs}

In this section, we analyze the observed line-of-sight (LOS) velocity
distributions for Sculptor (Scl) and Ursa Minor (UMi). Our goal is to
understand if the galaxies show evidence of tidal features---for
example, a velocity gradient or rising outer velocity dispersion. Our
derived systemic velocities and dispersions are consistent with
literature. We find weak evidence for a velocity gradient in Scl and a
rising outer velocity dispersion. However, Scl's velocity gradient is
misaligned with its proper motion, likely inconsistent with a tidal
interpretation. Scl's velocity dispersion also mostly rises within the
effective radius, more likely a feature of inner structure than tidal
disruption. We find no evidence of a gradient in mean velocity or
velocity dispersion in UMi. We conclude that Scl and UMi do not show
clear features of tidal disruption given current velocity observations.

\section{Data selection}\label{data-selection}

For both Sculptor and Ursa Minor, we construct literature samples of LOS
velocity measurements. We combine these samples with J+24's Bayesian
likelihoods to produce LOS velocity-informed membership probabilities.
We select stars with moderate to large membership probabilities for
future analysis.

\begin{table*}[t]
\centering
\caption[Spectroscopic LOS velocity measurements]{Summary of velocity measurements and derived properties. }
\label{tbl:rv_measurements}
\begin{tabular}{llllllll}
\toprule
 & Study & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$\\
\midrule
Scl & combined & 8945 & 2280 & 2096 & 1981 & 0.9 & \\
 & tolstoy+23 & 3311 & 1701 & 1522 & 1482 & 0.65 & –\\
 & sestito+23a & 2 & 2 & 2 & 2 & 13 & –\\
 & walker+09 & 1818 & 1522 & 1417 & 1328 & 1.8 & 3”\\
 & APOGEE & 5082 & 253 & 170 & 164 & 0.5 & –\\
UMi & combined & 4714 & 1225 & 1148 & 863 & 2.1 & \\
 & sestito+23b & 5 & 5 & 5 & 5 & 1.8 & –\\
 & pace+20 & 1716 & 1538 & 829 & 678 & 2.5 & 1”\\
 & spencer+18 & 1407 & 970 & 596 & 406 & 0.9 & 2”\\
 & APOGEE & 9500 & 279 & 37 & 67 & 0.6 & –\\
\bottomrule
\end{tabular}
\end{table*}

First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
not report GaiaDR3 source ID's, we match to the nearest star within 1--3
arcseconds (see REF Table~\ref{tbl:rv_measurements}). We combine the
mean RV measurement from each study using the inverse-variance weighted
mean \(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
variance \(s^2\).

We next remove stars with significant velocity dispersions as measured
between observations in a study or between studies, to filter out
unreliable measurements or possible spectroscopic binaries. By using
that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
distribution with \(N-1\) measurements. This cut typically removes stars
with reduced chi-squared values
\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (as a star
typically fewer than four velocity measurements).

We correct the velocities for the solar motion and on-sky size of the
galaxy. First, we transform the velocities into the galactic standard of
rest (GSR), a reference frame centred at the Sun's position but at rest
with respect to the Galactic centre. Next, we transform velocities to
correct for the apparent gradient induced by the dwarf's proper motion.
Let the \(\hat z\) be the direction from the sun to the dwarf galaxy.
Then if \(\phi\) is the angular distance between the centre of the
galaxy and the individual star, the corrected radial velocity is then
\begin{equation}{
v_{\rm gsr}' = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
}\end{equation} \{eq:v\_z\} where \(v_{\rm los, gsr}\) is the LOS
velocity in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the
tangental velocities in RA and Dec in the GSR frame, and \(\theta\) is
the position angle of the star with respect to the centre of the dwarf.
The correction from both effects induces an apparent gradient of about
\(1.3\,\kmsdeg\) for Sculptor and less for Ursa Minor \citep[see
also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
the distance uncertainty and velocity dispersion in quadrature to the RV
uncertainties for each star. We use the \(v_z\) values for the following
modelling.

We build on J+24's likelihood framework to add velocity information (see
Section~\ref{sec:extra_density}). The total likelihood of a star
belonging to the satellite or foreground/background population is
\begin{equation}{
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm los},
}\end{equation} where we have added \({\cal L}_{\rm los}\), LOS-velocity
likelihood term. We assume that the satellite and background
\(v_{\rm los, gsr}\) distributions are Gaussian. Specifically,
\begin{equation}{
\begin{split}
{\cal L}_{\rm los, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
{\cal L}_{\rm los, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
\end{split}
}\end{equation} where \(f\) is the probability density of a standard
normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
velocity and dispersion of the satellite, and \(\delta v_i\) is the
individual measurement uncertainty. Typically, the velocity dispersion
dominates the uncertainty budget. We assume a halo/background velocity
dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
\citep[e.g.][]{brown+2010}.

Similar to above, we retain stars with the resulting membership
probability of greater than 0.2. Like in J+24, most stars have
probabilities of nearly 1 or 0, so the precise choice of cut marginally
affects the resulting sample.

@@

To model for possible gradients. We use Monte-Carlo Markov chains (MCMC)
to calculate the posterior distributions of our models of the intrinsic
velocity dispersion and possible velocity gradients in Scl and UMi.

The likelihood is based on the distribution of stars with respect to the
model's predicted \(\mu_\V\) and \(\sigma_\V\) at each star's position
in the tangent plane \(\xi\) and \(\eta\). Specifically, the log
likelihood is \begin{equation}{
\log {\cal L} = \sum_i \frac{1}{\sigma_i} \log f\left(\frac{\V_i - \mu_i}{\sigma_i}\right),
}\end{equation} where \(f\) is the probability distribution of a
standard normal distribution, \(\V_i\) is the velocity of the \(i\)th
star, and \(\mu_i\) and \(\sigma_i\) are the model's predicted mean and
velocity dispersion at the \(i\)th star's position \(\xi_i\) and
\(\eta_i\).

We consider a single model designed to derive systemic velocity and the
velocity dispersion, detect velocity gradients, and detect a rising
velocity dispersion. The mean velocity at a given point on the sky,
\(\mu\), is assumed to be, \begin{equation}{
\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta
}\end{equation}

for tangent-plane coordinates \(\xi\) and \(\eta\), systemtic velocity
\(\mu_0\), and velocity gradient slopes \(a\) and \(b\). The velocity
dispersion at a given position, \(\sigma\), is assumed to depend as a
power-law on elliptical radius \(R_{\rm ell}\) alone: \begin{equation}{
\log \sigma = \log \sigma_0 + c\,\log(R_{\rm ell} / R_h)
}\end{equation} where \(\sigma_0\) is the system's velocity dispersion
at \(R_h\), and \(c\) is the velocity dispersion gradient slope.

Our model has parameters with priors \begin{equation}{
\mu_{0} \sim N(0, \sigma^2_{\rm halo}) \,\kms\\
\sigma_{0} \sim U(0, 20)\ \kms \\
a \sim N(0, 6^2)\ \kms\,{\rm deg}^{-1}\\
b \sim N(0, 6^2)\ \kms\,{\rm deg}^{-1}\\
c\sim N(0, 0.3^2)\ {\rm dex} \\
}\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
the velocity dispersion of the MW halo adopted above,
\(N(\mu, \sigma^2)\) is a standard normal distribution with mean \(\mu\)
and variance \(\sigma^2\), and \(U(l, h)\) is a uniform distribution
between \(l\) and \(h\).

Bayes factors quantify the relative evidence between Bayesian models. We
can use Bayes factors to compare the above model to simpler models
without gradients in \(\sigma\) or \(\mu\). Assuming that a
gradient-free model represents the case \(a=b=0\) (gradient free) or
\(c=0\) (fixed dispersion) models, we can quickly calculate the relative
Bayes factors without running another model. We use the Savage-Dickey
method \citep{dickey+lientz1970}, where the Bayes factor is then the
relative density of posterior versus prior samples when \(a=b=0\)
(gradient free) or \(c=0\) (fixed dispersion). In order to calculate
these densities, we use a kernel density estimator on the posterior
samples with a Silvermann bandwidth in each dimension.

While this is a more complex model, we find each parameter agrees with
models of reduced complexity.

@@

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/sigma_v_gradient.pdf}}
\caption[Possible gradients in the velocity dispersion]{The observed
velocity dispersion gradient (black) in 10 equal number bins and the
derived slopes from the model fitting. Scl shows moderate evidence for
an increasing velocity dispersion with radius. UMi's evidence for a
radial gradient is weaker, and Scl's velocity dispersion appears to be
flat outside of \(R_h\).}
\end{figure}

For Sculptor, we combine radial velocity measurements from APOGEE,
\citet{sestito+2023a}, \citet{tolstoy+2023}, and \citet{WMO2009}.
\citet{tolstoy+2023} and \citet{WMO2009} provide the bulk of the
measurements. We find that there is no significant velocity shift in
crossmatched stars between catalogues. After crossmatching to high
quality Gaia stars and excluding significant stellar velocity
dispersions, we have a sample of 1918 members.

We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
\citet{battaglia+2008}]{walker+2009}.

We detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at
a position angle of \(-149_{-13}^{+17}\) degrees. Several past work has
attempted to detect a gradient in Sculptor, but no consensus has been
reached. \citet{arroyo-polonio+2024} detect a velocity gradient of
\(4\pm1.5\,\kmsdeg\) in a similar direction using the
\citet{tolstoy+2023} sample, finding inconclusive statistical evidence.
They additionally suggest a third chemodynamical component of the galaxy
which may bias rotation measurements. \citet{battaglia+2008} also detect
a \(-7.6_{-2.2}^{+3.0}\,\kmsdeg\) velocity gradient along the major
axis, approximately the same direction. Instead, \citet{strigari2010};
\citet{martinez-garcia+2023} detect no significant gradient in Sculptor
using \citet{WMO2009} sample. Note that pre-\emph{Gaia} work did not
have as strong of a constraint on the proper motion of Scl, which limits
conclusions of the intrinsic velocity gradient in Scl.

Fig.~\ref{fig:scl_velocity_gradient_scatter} plots the combined velocity
sample and the MCMC samples for the velocity gradient. While some
samples have a consistent gradient with 0, most samples have a positive
velocity gradient, consistent with the rolling median trend.

The velocity gradient in Scl is misaligned with the proper motion

For UMi, we collect radial velocities from, APOGEE,
\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}. We
shifted the velocities of \citet{spencer+2018} (\(-1.1\,\kms\)) and
\citet{pace+2020} (\(+1.1\,\kms\) ) to reach the same scale. We found
183 crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
dispersion cut, and PSAT J+24 \textgreater{} 0.2 w/o velocities). Since
the median difference in velocities in this crossmatch is about 2.2
km/s, we adopt 1 km/s as the approximate systematic error here. Our
final sample includes 831 members.

We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
dispersion of \(8.8\pm0.2\,\kms\) for UMi. This is consistent with
\citet{pace+2020} and to a lesser extent with \citet{spencer+2018}. We
do not find evidence for a velocity gradient, consistent with past work
\citep{pace+2020, martinez-garcia+2023}.

For both Ursa Minor and Sculptor, we also fit models to only data from
individual surveys. Since the resulting parameters are very similar, we
conclude that many of the systematic uncertainties are likely smaller
than the present errors or that each large survey has similar biases.

\section{Discussion and limitations}\label{discussion-and-limitations}

Our model here is relatively simple. Some things which we note as
systematics or limitations:

\emph{Inter-study systematics and biases}. While basic crossmatches and
a simple velocity shift, combining data from multiple instruments is
challenging. This appears to be a minor issue (Sculptor) or is corrected
for (Ursa Minor).

\emph{Misrepresentative uncertainties}. Inspection of the variances
compared to the standard deviations within a study seems to imply that
errors are accurately reported. APOGEE notes that their RV uncertainties
are known to be underestimates but are a small proportion of our sample.
We find that most samples report internally-consistent uncertainties for
multiple observations of the same star.

\emph{Binarity}. Stars in binary systems can inflate the inferred
system's velocity dispersion. For example, a system with an apparent
velocity dispersion of \(9\,\kms\) may have a true velocity dispersion
of \(8\,\kms\) \citep{spencer+2017}. Thus, our measurement is likely
slightly inflated given the high binary fractions measured in these
systems \citep{spencer+2018, arroyo-polonio+2023, gration+2025}.

\emph{Multiple populations}. Both Sculptor and Ursa Minor likely contain
multiple populations \citep[\citet{pace+2020},
\citet{tolstoy+2004}]{arroyo-polonio+2024}. Since we only model a single
population, and each population may have a different extent and velocity
dispersion, this could result in biased velocity dispersions. However,
it is unclear how to uniquely determine an overall velocity dispersion
in a multi-population system.

\emph{Selection effects}. RV studies each have their own selection
effects, which may affect the resulting dispersion, especially if
different populations or regions of the galaxy have different velocities
or velocity dispersions. We do not attempt to correct for this. Given
that samples extend through the half-light radius of each system, we do
not expect current samples to miss large features except in the very
outskirts.

Could we observe velocity gradients instead in \emph{Gaia}.
Unfortunately, given Scl and UMi's distances (\(\gtrsim 70\,\kpc\))
\emph{Gaia's} systemic proper motion uncertainty of \(0.017\,\masyr\)
corresponds to a velocity of \(5\,\kms\), making the detection of
gradients of amplitude less than \(5\,\kms\) challenging. In addition,
the typical uncertainties of member stars of Scl and UMi are even
larger. While in future data releases, \emph{Gaia} uncertainties may
permit study in the internal motion of dwarfs, the current \emph{Gaia}
proper motions are not yet precise enough.

>>>>

\chapter{Line-of-sight velocities: Sample selection and
modelling}\label{sec:extra_rv_models}

In this section, we analyze the observed line-of-sight (LOS) velocity
distributions for Sculptor (Scl) and Ursa Minor (UMi). Our aim is to
test for kinematic tidal signatures by combining literature LOS
velocities with J+24's membership framework. Our derived systemic
velocities and dispersions are consistent with literature. We find weak
evidence for a velocity gradient in Scl. As the gradient is misaligned
with the orbit, the gradient more likely reflects intrinsic rotation
than tidal disruption. Scl's velocity dispersion also rises within the
effective radius, but predominantly in the inner regions, contrary to
tidal disruption. We find no evidence of a gradient in mean velocity or
velocity dispersion in UMi. We conclude that Scl and UMi do not show
clear features of tidal disruption given current velocity observations.

\section{Data processing and
selection}\label{data-processing-and-selection}

We compile LOS velocities from several spectroscopic surveys. For
Sculptor, we combine \citet{tolstoy+2023}; \citet{walker+2009};
\citet{sestito+2023a}; and \apogee{} \citep[DR17,][]{abdurrouf+2022}.
For Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
\citet{sestito+2023b}; and \apogee{}. We then crossmatch all catalogues
to J+24 Gaia stars. If a study did not report GaiaDR3 source ID's, we
match to the nearest star within 1--3 arcseconds. We combine
measurements of the same star using inverse-variance weighting.

To reduce likely binaries, we remove stars with significant velocity
dispersions.\footnote{Specifically, using that
  \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
  \(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
  distribution with \(N-1\) measurements.}

We build on J+24's likelihood by adding multiplicative terms in the
total likelihood for the velocity consistency (see
Section~\ref{sec:extra_density}). We assume that the satellite and
background \(v_{\rm los, gsr}\) distributions are Gaussian. We assume a
halo/background velocity dispersion of a constant
\(\sigma_{\rm halo} = 100\,\kms\) with mean 0
\citep[e.g.][]{brown+2010}. We select stars with velocity-informed
satellite membership probabilities of greater than 0.2. For Scl, we find
1918 unique members and UMi, 831.

For UMi, we shifted the velocities of \citet{spencer+2018}
(\(-1.1\,\kms\)) and \citet{pace+2020} (\(+1.1\,\kms\) ) to account for
a systematic velocity offset. Otherwise all studies appear to agree.

We correct the velocities for the solar motion and on-sky size of the
galaxy. We transform the velocities into the galactic standard of rest
(GSR, i.e.~same location as ICRS but velocities relative to the galactic
centre), and then transform velocities to correct for the apparent
gradient induced by the dwarf's proper motion.\footnote{Specifically,
  sub} We define \(v_{\rm gsr}'\) to be velocities in the The correction
from both effects induces an apparent gradient of about \(1.3\,\kmsdeg\)
for Sculptor and less for Ursa Minor \citep[see
also][]{WMO2008, strigari2010}. The uncertainty on this velocity
correction are less than the individual star uncertainties and our
derived velocity gradient.

@@

We fit Monte-Carlo Markov chain (MCMC) models to solve for the systemic
velocity, velocity dispersion, and possible gradient. We assume that the
galaxy follows a planar velocity gradient. The mean velocity at a given
point on the sky, \(\mu\), is assumed to be,

\begin{equation}{
\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta
}\end{equation}

for tangent-plane coordinates \(\xi\) and \(\eta\), systemic velocity
\(\mu_0\), and velocity gradient slopes \(a\) and \(b\). The velocity
dispersion at a given position, \(\sigma\), is assumed to depend as a
power-law on elliptical radius \(R_{\rm ell}\) alone: \begin{equation}{
\log \sigma = \log \sigma_0 + c\,\log(R_{\rm ell} / R_h)
}\end{equation} where \(\sigma_0\) is the system's velocity dispersion
at \(R_h\), and \(c\) is the velocity dispersion gradient slope. We use
weakly-informative priors, as described in Table~\ref{tbl:scl_rv_mcmc},
and with \(a, b \sim N(0, 6^2)\,\kmsdeg\).

@@

Figure: The observed velocity dispersion gradient (black) in 10 equal
number bins and the derived slopes from the model fitting. Scl shows
moderate evidence for an increasing velocity dispersion with radius.
UMi's evidence for a radial gradient is weaker, and Scl's velocity
dispersion appears to be flat outside of \(R_h\).

@@

We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
\citet{battaglia+2008}]{walker+2009}. We detect a moderately significant
gradient of \(4.3\pm1.3\,\kmsdeg\) at a position angle of
\(-149_{-13}^{+17}\) degrees, similar to past detections
\citetext{\citealp[e.g.,][]{arroyo-polonio+2024}; \citealp{battaglia+2008}; \citealp[but
see also][]{strigari2010}; \citealp{martinez-garcia+2023}}.

Fig.~\ref{fig:scl_velocity_gradient_scatter} plots the combined velocity
sample and the MCMC samples for the velocity gradient. While some
samples have a consistent gradient with 0, most samples have a positive
velocity gradient, consistent with the rolling median trend. The
velocity gradient in Scl is misaligned with the proper motion

We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
dispersion of \(8.8\pm0.2\,\kms\) for UMi. We do not find evidence for a
velocity gradient. This is consistent with previous work
\citetext{\citealp{pace+2020}; \citealp[somewhat
with][]{spencer+2018}; \citealp{martinez-garcia+2023}}.

\section{Discussion and caveats}\label{discussion-and-caveats}

\emph{Binarity}. Stars in binary systems can inflate the inferred
system's velocity dispersion. For Scl and UMi with high measured
velocity dispersion, binaries likely add \(\sim 1\,\kms\) to the
velocity dispersion \citep{spencer+2018, gration+2025}.

\emph{Multiple populations}. Both Sculptor and Ursa Minor likely contain
multiple populations with different kinematics\citep[\citet{pace+2020},
\citet{tolstoy+2004}]{arroyo-polonio+2024}. We do not model these
separately---it is unclear how to define a velocity dispersion in such a
case.

\emph{Inter-study biases}. While basic crossmatches and a simple
velocity shift, combining data from multiple instruments is challenging.
Studies may also have biased selection effects or misrepresentative
uncertainties, biasing our results. However, besides the \(\sim1\,\kms\)
velocity offset in Ursa Minor, our results are consistent across
samples.

\emph{Tangental motions.} Precise star-by-star proper motions would
further test for kinematic disequilibrium. Presently, \emph{Gaia}
uncertainties are too large to permit such studies for Sculptor and Ursa
Minor.\footnote{Specifically, typical proper motion uncertntanties for
  faint, member stars of Scl and UMi are around \(0.5\,\masyr\),
  corresponding to velocities \(\sim 100\,\kms\).}
